---
title: "Abortion Clinic Analysis"
author: "Sam Hart"
date: "5/6/2022"
output: 
  html_document:
    keep_md: yes
  md_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(formattable)
library(scales)

(counties <- read_csv('../mid/distance_census.csv'))
```


```{r distance, include=FALSE}

avg_distance <- mean(counties$miles)
avg_distance_roe <- mean(counties$miles_roe)

counties <- counties %>%
  mutate(distance_bucket = floor(miles / 100) * 100)

really_far <- counties %>%
  filter(distance_bucket >= 500)

far_women_likely <- counties %>%
  filter(miles >= 300) 

far_women_certain <- counties %>%
  filter(miles_strict >= 300) 

far_women_now <- counties %>%
  filter(miles_roe >= 300) 

women_far_likely <- sum(far_women_likely$B01001026)
women_far_certain <- sum(far_women_certain$B01001026)
women_far_now <- sum(far_women_now$B01001026)


```
## How many women live far from an abortion clinic? 

With Roe in place, **`r comma(women_far_now)`** women live more than 300 mi from an abortion clinic. 

If Roe is overturned, it's almost certain that **`r comma(women_far_certain)`** will live more than 300 mi from the nearest abortion clinic operating without restrictions.

It is possible and likely that number could be closer to  **`r comma(women_far_likely)`** if certain states impose restrictions.

## Average distance to nearest clinic

This is how things look now, with every clinic's operations protected by the Roe v. Wade decision.

The average distance to the nearest clinic for each county is **`r avg_distance_roe` mi**.


```{r, map1, echo=FALSE}
library(ggplot2)
library(usmap)

#add 'fips' column to be read by plot_usmap
distance_df <- counties %>%
  mutate(fips = str_replace(geoid,'05000US','')) %>%
  select(miles, miles_roe, miles_strict, fips)

plot_usmap(regions= "counties", data=distance_df, values='miles_roe', color='#ffffff', size=0) +
  labs(title="distance by county to the nearest clinic (without restrictions)") +
  scale_fill_continuous(type='viridis', direction=-1, option='magma', limits=c(0, 500))
```
If Roe v. Wade is overturned, certain clinics may be shut down by laws in states that choose to do so.

For each county, distance (mi) was computed to the nearest abortion clinic in a state without trigger laws or restrictions.

Average distance is **`r avg_distance` mi** to the nearest clinic if Roe is overturned.
``` {r, map2, echo=FALSE}
plot_usmap(regions= "counties", data=distance_df, values='miles_strict', color='#ffffff', size=0) +
  labs(title="distance by county to the nearest clinic if roe is overturned (almost certain)") +
  scale_fill_continuous(type='viridis', direction=-1, option='magma', limits=c(0, 300))
```
```{r, echo=FALSE}
plot_usmap(regions= "counties", data=distance_df, values='miles', color='#ffffff', size=0) +
  labs(title="distance by county to the nearest clinic if roe is overturned (likely)") +
  scale_fill_continuous(type='viridis', direction=-1, option='magma', limits=c(0, 300))
```

```{r}


avg_women_in_poverty <- summarise_if(counties[c('pct_women_in_poverty')], is.numeric, mean, na.rm = TRUE) * 100

avg_women_in_poverty <- avg_women_in_poverty[[1]]

## old buckets: discrete miles aka 100 200 300 400 500
poverty_buckets <- counties %>%
  filter(!is.na(pct_women_in_poverty)) %>%
  group_by(distance_bucket) %>%
  summarise(mean_pov = mean(pct_women_in_poverty) * 100, n = n()) %>%
  rename('Avg. % Women in poverty' = mean_pov, 'Distance from nearest clinic (mi)' = distance_bucket)

## quintile calculations
q = c(0.2, 0.4, 0.6, 0.8)
pov_by_mile_quant <- counties %>%
  filter(!is.na(pct_women_in_poverty))

mile_quants <- quantile(counties$miles, q)

pov_by_mile_quant <- pov_by_mile_quant %>%
  mutate(mile_quant = findInterval(miles, mile_quants)) %>%
  group_by(mile_quant) %>%
  summarise(mean_pov = mean(pct_women_in_poverty) * 100, n = n()) %>%
  mutate(mile_quant = (mile_quant + 1) * 20) %>%
  rename('% women in poverty' = mean_pov, 'quintile of distance from nearest clinic (nth)' = mile_quant)
  
```
## Demographics

Average % women in poverty in all counties: **`r avg_women_in_poverty `**

#### % Women in poverty by distance from nearest Post-Roe clinic
```{r echo=FALSE, results='asis'}
library(knitr)
kable(pov_by_mile_quant, 'pipe')
```

```{r, map3, echo=FALSE}
#add 'fips' column to be read by plot_usmap
poverty_df <- counties %>%
  mutate(fips = str_replace(geoid,'05000US','')) %>%
  mutate(pov_x_distance = pct_women_in_poverty * miles) %>%
  select(pov_x_distance, fips)

plot_usmap(regions= "counties", data=poverty_df, values='pov_x_distance', color='#ffffff', size=0) +
  scale_fill_continuous(type='viridis', direction=-1, option='magma')
```

```{r pressure, echo=FALSE}

avg_women_wo_ins <- summarise_if(counties[c('pct_women_wo_insurance')], is.numeric, mean, na.rm = TRUE) * 100

avg_women_wo_ins <- avg_women_wo_ins[[1]]

insurance_buckets <- counties %>%
  filter(!is.na(pct_women_wo_insurance)) %>%
  group_by(distance_bucket) %>%
  summarise(mean_ins = mean(pct_women_wo_insurance) * 100, n = n()) %>%
  rename('Avg. % Women without insurance' = mean_ins, 'Distance from nearest clinic (mi)' = distance_bucket)
ins_by_mile_quant <- counties %>%
  filter(!is.na(pct_women_wo_insurance))

ins_by_mile_quant <- ins_by_mile_quant %>%
  mutate(mile_quant = findInterval(miles, mile_quants)) %>%
  group_by(mile_quant) %>%
  summarise(mean_ins = mean(pct_women_wo_insurance) * 100, n = n()) %>%
  mutate(mile_quant = mile_quant + 1) %>%
  rename('% women without health insurance' = mean_ins, 'quintile of distance from nearest clinic (nth)' = mile_quant)

```

Average % women without insurance in all counties: **`r avg_women_wo_ins `**

#### % Women without insurance by distance from nearest Post-Roe clinic
```{r echo=FALSE, results='asis'}
library(knitr)
kable(ins_by_mile_quant, 'pipe')
```

```{r, map4, echo=FALSE}
#add 'fips' column to be read by plot_usmap
insurance_df <- counties %>%
  mutate(fips = str_replace(geoid,'05000US','')) %>%
  mutate(no_ins_x_distance = pct_women_wo_insurance * miles) %>%
  select(no_ins_x_distance, fips)

plot_usmap(regions= "counties", data=insurance_df, values='no_ins_x_distance', color='#ffffff', size=0) +
  scale_fill_continuous(type='viridis', direction=-1, option='magma')
```

